Viscous Erosion in Porous Media

### 2018-02-19 docs/methods/erosion_methods.tex
Finally wrote the time-stepping section in a way that I am happy with. It is more elegant with applying RK2 directly to the argument of the integrating factor.

In addition to the curvature-dependent term in~\eqref{eqn:Vn}, we
regularize boundary evolution by applying a narrow Gaussian filter to
absolute shear stress
\begin{align*}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma} \exp \left(-
\frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \,
d\alpha'.
\end{align*}
Here $\sigma$ is a small parameter that determines the width of the
filter. The corresponding normal and tangential interface velocities are
the given by
\begin{align*}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align*}

For convenience, we introduce the variables
\begin{align*}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atausig}}{L}.
\end{align*}
Then, the evolution equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal terms) in
the PDE for $\theta$.

We can evolve $\theta$ in spectral space since $\theta(\alpha)$ is
periodic up to a jump of $2\pi$ for $\alpha \in [0,2\pi]$.  Therefore,
we introduce the Fourier series
\begin{align}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that we have separated out the linear
part of $\theta$, so that the remaining function is smooth and has a
rapidly decaying Fourier series.  In spectral
space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
Each ODE is first-order and linear (though variable-coefficient due to
$\elfun$) and thus can be solved by using the integrating factor
\begin{align}
  \label{eq:mu}
  \mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k.
\end{align}
We solve system~\eqref{eqn:Levo3}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we
choose the {\em midpoint} RK2 method so as to maximize the smoothing
effect of the curvature term.

At the first stage of RK2 (the half step), we discretize
equations~\eqref{eqn:Levo3}--\eqref{eqn:thetaevo3} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have
chosen the constant of integration in the exponent of
equation~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of
RK2, we must compute $\Mterm^{(n+1/2)}$ and
$\widehat{\NLterm}_k^{(n+1/2)}$, which requires a call to the Stokes
solver. Hence, we must compute $\mu_k^{(n+1/2)}$, so that
$\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step
can be computed. To compute $\mu_k^{(n+1/2)}$, we use the left-end-point
rule (first order, consistent with first forward Euler step of RK2)
\begin{align*}
\mu_k^{(n+1/2)} = \exp \left( \epsilon k^2 \frac{\Dt}{2}
\elfun^{(n)}\right).
\end{align*}
It is instructive to recast as
\begin{align*}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{align*}
so that the smoothing effect of the exponential is obvious. With
$\thhat_k^{(n+1/2)}$ computed, it is now possible to compute
$\Mterm^{(n+1/2)}$ and $\widehat{\NLterm}_k^{(n+1/2)}$.

Next, at the second stage of RK2, we discretize as
\begin{align*}
\frac{L^{(n+1)} - L^{(n)}}{\Dt} &= \Mterm^{(n+1/2)}, \\
\frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} &=
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align*}

We compute the integrating factor with RK2 to get
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
& \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}

%We again compute the integrating factor, $\mu_k^{(n+1)}$ via the
%\todo[inline]{insert integration rule}, giving
%\begin{align*}
%\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
%& \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{4} \eps k^2 \Dt \left( 3 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
%\end{align*}

>>>>

In addition to the curvature-dependent term in~\eqref{eqn:Vn}, we regularize boundary evolution by applying a narrow Gaussian filter to absolute shear stress
\begin{align*}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma} \exp \left(-
\frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \,
d\alpha'.
\end{align*}
Here $\sigma$ sets the width of the filter. The corresponding normal and tangential interface velocities are then given by
\begin{align*}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align*}

For convenience, we introduce the variables
\begin{align*}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align*}
Then, the evolution equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal terms) in
the PDE for $\theta$.

We will evolve $\theta(\alpha)$ spectrally and so introduce the Fourier series
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that in \eqref{thetaSeries}, we have decomposed $\theta(\alpha)$ into a linear component and a periodic component.
In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
Each ODE is first-order and linear (though variable-coefficient due to
$\elfun$) and thus can be solved by using the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1
\end{align}
We solve system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step) we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as
\begin{align}
\label{eq:mu12d}
& \mu_k^{(n+1/2)} = \mu_k^{(n)} \exp \left( \eps k^2 \frac{\Dt}{2} \zeta^{(n)} \right) \\
\label{eq:L12d}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
\label{eq:th12d}
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt/2} 
= \mu_k^{(n)} \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. At the next stage of RK2 it will be necessary to evaluate derivates at the half step $n+1/2$, thus requiring a call to the Stokes solver for which the shape of the bodies must be known through $\theta^{(n+1/2)}$. We therefore use \eqref{eq:mu12d} to solve \eqref{eq:th12d} for $\theta^{(n+1/2)}$, giving
\begin{align}
\label{eq:th12}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{align}
Since $\zeta >0$, the exponential term always acts to smooth the geometry and thus mitigate numerical instabilities. With the geometry at step $n+1/2$ known, a call to the Stokes solver allows one to compute the terms $\Mterm^{(n+1/2)}$, $\widehat{\NLterm}_k^{(n+1/2)}$, and $\zeta^{(n+1/2)}$.

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we use \eqref{eq:mu1d} to solve \eqref{eq:th1d} for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to reduce high-frequency oscillations in shape.
